Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-03 manuscript/manuscript.tex
edits to sky scramble section

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}

Figure~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORF, respectively.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
for a particular data set to the 
distribution of noise-marginalized optimal statistics obtained using sky scrambles.
\sv{sky scrambles for all of our simulated data sets}

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/skyscrambles_dataset50.pdf}
	\caption{Comparison between the optimal statistic with and without sky scrambles 
			for a simulated dataset containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			The true analysis found an $\mathrm{SNR} = 1.1$, while only 24 of the 210 sky scramble analyses 
			found $\mathrm{SNR} \geq 1.1$ ($p=0.11$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

\sv{comparison to Steve's}

>>>>

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background. 
We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
for a common red process. For this particular realization of the stochastic background, 
the values of $\hat{A}^2$ from the noise-marginalized and Bayesian analyses are in good agreement 
with each other and the injected GW stochastic background, 
but the value from the fixed-noise analysis is significantly lower. 
The noise-marginalized analysis also gives a larger mean SNR compared to the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (solid orange line). 
			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
			for a common red process (solid green line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
	\label{fig:os_dataset_sample}
\end{figure}

Figure~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORF, respectively.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

Figure~\ref{fig:skyscrambles_dataset_sample} compares the noise-marginalized optimal statistic 
for a particular data set to the distribution of noise-marginalized optimal statistics obtained using sky scrambles. 
For this data set, the noise-marginalized mean optimal statistic is 
$1.9\times10^{-29}$ and the mean SNR is 1.1. 
The distributions of the noise-marginalized optimal statistic and SNR obtained using the scrambled ORFs 
are both centered around zero, 
and for 24 out of the 210 sky scrambles, the SNR is above the true value of $1.1$ ($p=0.11$). 
Figure~\ref{fig:skyscrambles} shows the $p$-values for 210 sky scrambles 
using 425 realizations of the stochastic background. 
For 62\% of simulations the $p$-value is $<0.1$, and for 29\% of simulations the $p$-value is $<0.01$.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/skyscrambles_dataset50.pdf}
	\caption{Comparison between the optimal statistic with and without sky scrambles 
			for a simulated dataset containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			The true analysis found an $\mathrm{SNR} = 1.1$, while only 24 of the 210 sky scramble analyses 
			found $\mathrm{SNR} \geq 1.1$ ($p=0.11$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_skyscrambles.pdf}
	\caption{Distribution of $p$-values for 425 different realizations of the stochastic background. 
			For each realization, the $p$-value is the fraction of sky scrambled ORFs that gives a 
			noise-marginalized mean SNR greater than the mean SNR using the correct ORF.}
	\label{fig:skyscrambles}
\end{figure}

For five realizations of the stochastic background, we also computed the Bayes' factors for the 
true and sky scrambled ORFs in order to directly compare 
the frequentist and Bayesian detection techniques.
\sv{comparison to Steve's}
